Stochastic Variance Reduced Policy Gradient

### 2018-02-01 svrpg.tex

\section{Preliminaries}
In this section we provide essential background on policy gradient methods for Reinforcement Learning and stochastic variance-reduced gradient methods for finite-sum optimization. In the next section we will show how these methods can be combined to solve Reinforcement Learning tasks.

\subsection{Policy Gradient}\label{subsec:PolicyGradient}
A Reinforcement Learning task \cite{sutton2011reinforcement} can be modeled with a discrete-time continuous Markov Decision Process (MDP) $\langle\Sspace,\Aspace,\Transition,\Reward,\gamma,\rho\rangle$, where $\Sspace$ is a continuous state space; $\Aspace$ is a continuous action space; $\Transition$ is a Markovian transition model, where $\Transition(s'|s,a)$ defines the transition density from state $s$ to $s'$ under action $a$; $\Reward$ is the reward function, where $\Reward(s,a) \in [-R,R]$ is the expected reward for state-action pair $(s,a)$ and $R$ is the maximum absolute-value reward; $\gamma\in[0,1)$ is the discount factor; and $\rho$ is the initial state distribution.
The agent's behavior is modeled as a policy $\pi$, where $\pi(\cdot|s)$ is the density distribution over $\Aspace$ in state $s$.
% We consider episodic tasks, \ie tasks composed of episodes of length $H$, also called time horizon.
% In this setting, a trajectory $\tau$ is the sequence of states and actions $(s_0,a_0,s_1,a_1,\dots,s_{H-1},a_{H-1})$ observed by following some stationary policy in a given episode, where $s_0$ is always sampled from the initial state distribution $\rho$.
We consider episodic MDPs with effective horizon $H$, \ie optimal policy is able to reach the target state (\ie absorbing state) in at most $H$ steps (the true length depend on the initial state). This has not to be confused with a finite horizon problem where the optimal policy is non-stationary. In this setting, we can limit our attention to trajectories of length $H$. A trajectory $\tau$ is a sequence of states and actions $(s_0,a_0,s_1,a_1,\dots,s_{H-1},a_{H-1})$ observed by following a stationary policy, where $s_0 \sim \rho$.
With abuse of notation, we denote with $\pi(\tau)$ the density distribution induced by policy $\pi$ on the set $\Tspace$ of all possible trajectories, and with $\Reward(\tau)$ the total discounted reward provided by trajectory $\tau$:
\begin{align*}
\Reward(\tau) = \sum_{t=0}^{H-1}\gamma^t\Reward(s_t,a_t).
\end{align*}
Policies can be ranked based on their expected total reward:
\begin{align*}
	J(\pi) = \int_{\Tspace}\pi(\tau)\Reward(\tau)\de \tau.
\end{align*}
Solving an MDP means finding $\pi^* \in \argmax_{\pi} \{J(\pi)\}$.

Policy gradient methods restrict the search for the performance-maximizing policy over a class of parametrized policies $\Pi_{\vtheta}=\{\pol: \vtheta \in \realspace^m\}$, with the only constraint that $\pol$ is differentiable \wrt the $m$-dimensional parameter vector $\vtheta$. We denote the performance of a parametric policy with $J(\vtheta)$ for brevity, while in some occasions we replace $\pol(\tau)$ with $\pi(\tau|\vtheta)$ for the sake of readability. The search for a locally optimal policy is performed by means of gradient ascent, where the gradient of $J(\vtheta)$ \wrt the policy parameters is \cite{sutton2000policy}:
\todopir{stochastic or mini-batch gradient}
\begin{align*}
\gradJ{\vtheta} = \int_{\Tspace}\pol(\tau)\score{\vtheta}{\tau}\Reward(\tau)\de \tau.
\end{align*}
We consider the on-line learning problem, where trajectories must be sampled directly by interacting with the environment. 
In this setting, $(N>0)$-batch gradient ascent is typically employed: after sampling a batch of $N$ trajectories (\ie $\Dataset_{N}$),
policy parameters are updated as $\vtheta\gets\vtheta + \alpha\gradApp{\vtheta}{N}$, where $\alpha$ is a step size and $\gradApp{\vtheta}{N}$ is an estimate of the true gradient. 
\todopir{step size is scalar or vectorial (just for notation)? Check consistency of $N$ with finite-sum function, we can use even $b$ or $B$ for the batch size.}
The REINFORCE gradient estimator~\citep{williams1992simple} provides a simple, unbiased way of estimating the gradient:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}\left(\sum_{h=0}^{H-1}\score{\vtheta}{a_h^n\mid s_h^n}\right)\left(\sum_{h=0}^{H-1}\gamma^h r_h^n - b\right),
\end{align*}
where subscripts denote the time step, superscripts denote the trajectory, $r_h$ is the reward actually collected at time $h$ and b can be any baseline, provided it is constant \wrt to actions.
\todopir{Actually it can be state-action dependant~\citep[\eg][]{Thomas2017actionbaseline,anonymous2018variance}.}
The G(PO)MDP gradient estimator~\cite{baxter2001infinite} is a refinement of REINFORCE which is subject to less variance while preserving the unbiasedness:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}\sum_{h=0}^{H-1}\left(\sum_{k=0}^{h}\score{\vtheta}{a_k^n\mid s_k^n}\right)\left(\gamma^h r_h^n - b\right).
\end{align*}
\todopir{Cite Jan's paper somewhere}
\todopir{At this point we need to connect with the supervised learning framework (we can just define everything inside the first sum as $\nabla g(\tau)$).}

@@

\section{Algorithm}\label{sec:algo}

@@

\begin{enumerate}
	\item \textit{Non-convexity}: the objective $J(\vtheta)$ is typically non-convex;
	\item \textit{Approximation}: the dataset $\mathcal{T}$ can be infinite, but we can only sample a finite number of trajectories;
	\item \textit{Non-stationarity}: the value of the parameter $\vtheta$ influences the sampling of trajectories;
\end{enumerate}

@@

\section{Convergence Guarantees}
In this section we adapt existing convergence guarantees for nonconvex SVRG \cite{reddi2016stochastic} \cite{allen2016variance} to the Policy Gradient framework.
Each of the three challenges presented at the beginning of Section \ref{sec:algo} can potentially prevent convergence, so we need additional assumptions:
\begin{enumerate}
\item \textit{Non-convexity}:
	as already mentioned, a common assumption for nonconvex optimization is the L-smoothness of the objective. We actually require the following, which is stronger:
	\begin{assumption}\label{ass:bounded_score}
		For each trajectory $\tau$, value of $\vtheta$ and dimensions $i,j$ there are positive constants $G,F<\infty$ such that:
		\begin{align*}
		&\left|\nabla_{\theta_i}\log\pi_{\vtheta}(\tau)\right| \leq \GRADLOG \\
		&\left|\frac{\partial^2}{\partial\theta_i\partial\theta_j}\log\pi_{\vtheta}(\tau)\right| \leq \HESSLOG
		\end{align*}
	\end{assumption}
	This property is satisfied by the commonly used Gaussian policy, and is a sufficient condition for the L-smoothness of the objective (see Lemma \ref{lemma:lsmooth} in the appendix).
\item \textit{Approximation}:
since we can only compute an estimate of the full gradient, we require its variance to be bounded:
	\begin{assumption}\label{ass:REINFORCE}
		There is a constant $V<\infty$ such that, for each policy $\pol$:
		\[
		\Tr\mathbb{V}ar\left(\gradApp{\vtheta}{1}\right) \leq \VARRF.
		\]
	\end{assumption}
This assumption is satisfied by the REINFORCE and G(PO)MDP estimators and Gaussian policies \cite{zhao2011analysis} \cite{pirotta2013adaptive}. The problem of full gradient approximation for SVRG has been addressed in the supervised learning framework \cite{harikandeh2015stopwasting} with results similar to ours.
\item \textit{Non-stationarity}:
the variance introduced by importance sampling must be bounded:
	\begin{assumption}\label{ass:M2}
		There is a constant $W<\infty$ such that, for each pair of policies encountered in Algorithm \ref{alg:svrpg} and for each possible trajectory,
		\[
		\mathbb{V}ar\left(\omega(\tau)\right) \leq \VARIS.
		\]
	\end{assumption}
Differently from the previous two, this assumption must be enforced by a proper handling of the epoch size $m$.
\end{enumerate}

We can now state the following key result:
\begin{restatable}[Convergence of the SVRPG algorithm]{theorem}{convergence}\label{theo:convergence}
Under Assumptions \ref{ass:bounded_score}, \ref{ass:REINFORCE} and \ref{ass:M2}, the parameter vector $\vtheta_A$ returned by Algorithm \ref{alg:svrpg} after $T=m\times S$ iterations has, for some positive constants $\psi,\zeta, \xi$ and for proper choice of the step size $\alpha$ and the epoch size $m$, the following property:
\begin{align*}
	&\EVV[]
	{\norm[]{\nabla J(\vtheta_A)}^2} 
		\leq
		\frac{J(\vtheta^*)-J(\vtheta_0)}{\psi T} +
		\frac{\zeta}{N}
		+\frac{\xi}{B}
\end{align*}
\end{restatable}
Specific constants and meta-parameter constraints are provided in the proof of the theorem in Appendix \ref{app:proofs}. In Section \ref{sec:stopping}, instead, we propose a joint selection of step size $\alpha$ and epoch size $m$ which is more suited for applications. 
The $\mathcal{O}(\nicefrac{1}{T})$ term is coherent with results on non-convex SVRG, \eg \cite{reddi2016stochastic}; the $\mathcal{O}(\nicefrac{1}{N})$ term is due to full gradient approximation and is analogous to the one in \cite{harikandeh2015stopwasting}; the $\mathcal{O}(\nicefrac{1}{B})$ term is due to importance weighting. To achieve true convergence, both the batch size $N$ and the mini-batch size $B$ should increase over time. In practice, it is enough to choose $N$ and $B$ large enough to make the second and the third term negligible, \ie to mitigate the variance introduced by full gradient approximation and importance sampling respectively. Since the optimal configuration may be task-dependent, we leave this choice to the empirical analysis of Section \ref{sec:exp}.

@@

\section{Experiments}\label{sec:exp}
We have designed our experiments to investigate whether:
\begin{enumerate}
\item The algorithm reaches high value of the reward function faster than GPOMDP.
\item Averagely SVRPG keep the reward function more stable than GPOMDP.
\end{enumerate}

@@

\subsection*{Ancillary Lemmas}
Before addressing the main convergence theorem, we prove some useful lemmas.

\begin{restatable}[]{lemma}{L-smoothness}\label{lemma:lsmooth}
	 Under Assumption \ref{ass:bounded_score}, $J(\vtheta)$ is L-smooth.
\end{restatable}
\begin{proof}
\todomat{...}
\end{proof}

@@

\section{Applicability to Gaussian Policies}\label{app:pol}
\todomat{...}

\section{Experimental Details}\label{app:exp}
Setting specifications for all the used environments:
\begin{enumerate}
	\item \emph{Cart-Pole Balancing} : 4-dimensional state space: cart position x, pole angle $\theta$, cart velocity $\dot{x}$ and pole velocity $\dot{\theta}$; 1-dimensional action space: the horizontal force applied to the cart body. Reward function  is defined as $r(s, a) := 10 - (1 - cos(\theta)) - 10^{-5}\norm[] a^2$. The episodes terminates when $|x|>2.4$ or $|\theta|>0.2$ or the number of time steps T is greater than 100.
	\item \emph{Mujoco Swimmer}: 12-dimensional state space: 3 links velocities ($v_x$ and $v_y$ of center of masses) and 2 actuated joints angles. 2-dimensional action space: the two momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 10^{-4}\norm[2] a^2$. The episodes terminates when the number of time steps T is greater than 500.
	\item \emph{Mujoco Half Cheetah}: 20-dimensional state space: 9 links and 6 actuated joints angles. 6-dimensional action space: the 6 momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 0.05\norm[2] a^2$. The episodes terminates when the number of time steps T is greater than 500.
\end{enumerate}

>>>>

\section{Preliminaries}
In this section we provide the essential background on policy gradient methods for Reinforcement Learning and stochastic variance-reduced gradient methods for finite-sum optimization.
%In the next section we will show how these methods can be combined to solve Reinforcement Learning tasks.

\subsection{Policy Gradient}\label{subsec:PolicyGradient}
A Reinforcement Learning task \cite{sutton2011reinforcement} can be modeled with a discrete-time continuous Markov Decision Process (MDP) $\langle\Sspace,\Aspace,\Transition,\Reward,\gamma,\rho\rangle$, where $\Sspace$ is a continuous state space; $\Aspace$ is a continuous action space; $\Transition$ is a Markovian transition model, where $\Transition(s'|s,a)$ defines the transition density from state $s$ to $s'$ under action $a$; $\Reward$ is the reward function, where $\Reward(s,a) \in [-R,R]$ is the expected reward for state-action pair $(s,a)$;
% and $R$ is the maximum absolute-value reward;
$\gamma\in[0,1)$ is the discount factor; and $\rho$ is the initial state distribution.
The agent's behavior is modeled as a policy $\pi$, where $\pi(\cdot|s)$ is the density distribution over $\Aspace$ in state $s$.
% We consider episodic tasks, \ie tasks composed of episodes of length $H$, also called time horizon.
% In this setting, a trajectory $\tau$ is the sequence of states and actions $(s_0,a_0,s_1,a_1,\dots,s_{H-1},a_{H-1})$ observed by following some stationary policy in a given episode, where $s_0$ is always sampled from the initial state distribution $\rho$.
We consider episodic MDPs with effective horizon $H$, \ie the optimal policy is able to reach the target state (\ie absorbing state) in at most $H$ steps (the true length depend on the initial state). This has not to be confused with a finite horizon problem where the optimal policy is non-stationary. In this setting, we can limit our attention to trajectories of length $H$. A trajectory $\tau$ is a sequence of states and actions $(s_0,a_0,s_1,a_1,\dots,s_{H-1},a_{H-1})$ observed by following a stationary policy, where $s_0 \sim \rho$.
With abuse of notation, we denote with $\pi(\tau)$ the density distribution induced by policy $\pi$ on the set $\Tspace$ of all possible trajectories, and with $\Reward(\tau)$ the total discounted reward provided by trajectory $\tau$:
%
$\Reward(\tau) = \sum_{t=0}^{H-1}\gamma^t\Reward(s_t,a_t).$
%
Policies can be ranked based on their expected total reward:
\begin{align*}
	J(\pi) = \int_{\Tspace}\pi(\tau)\Reward(\tau)\de \tau.
\end{align*}
Solving an MDP $M$ means finding $\pi^* \in \argmax_{\pi} \{J(\pi)\}$.

Policy gradient methods restrict the search for the performance-maximizing policy over a class of parametrized policies $\Pi_{\vtheta}=\{\pol: \vtheta \in \realspace^m\}$, with the only constraint that $\pol$ is differentiable \wrt $\vtheta$. We denote the performance of a parametric policy with $J(\vtheta)$ for brevity, while in some occasions we replace $\pol(\tau)$ with $\pi(\tau|\vtheta)$ for the sake of readability. The search for a locally optimal policy is performed by means of gradient ascent, where the gradient of $J(\vtheta)$ \wrt the policy parameters is \cite{sutton2000policy, Peters2008reinf}:
\begin{align} \label{E:policygradient}
\gradJ{\vtheta} = \int_{\Tspace}\pol(\tau)\score{\vtheta}{\tau}\Reward(\tau)\de \tau.
\end{align}
Notice that the distribution defining the gradient is induced by the current policy. This aspect introduces a nonstationarity in the sampling process. Since the underlying distribution is changing over time, it is necessary to resample at each update or use weighting techniques such as importance sampling.

We consider the \emph{on-line learning problem}, where trajectories are sampled by interacting with the environment at each policy change. 
In this setting, stochastic gradient ascent is typically employed.
At each iteration $k >0$, after sampling a batch of $N$ ($0<N<+\infty$) trajectories with policy $\pi_{\vtheta_k}$,
the policy is updated as $\vtheta  = \vtheta + \alpha\gradApp{\vtheta}{N}$, where $\alpha$ is a step size and $\gradApp{\vtheta}{N}$ is an estimate of~\eqref{E:policygradient} using $N$ samples.
Several estimators have been provided in literature as empirical average over trajectories:
\begin{align} \label{E:policygradient.estimate}
        \gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N} \nabla f(\tau^i).
\end{align}
The most widespread estimators are REINFORCE~\citep{williams1992simple} and G(PO)MDP~\citep{baxter2001infinite}.
We refer the reader to App.~\ref{A:gradient_estimators} for a formal definition.
Both are unbiased estimate of~\eqref{E:policygradient} but G(PO)MDP is subject to less variance than REINFORCE.
\todopir{At this point we need to connect with the supervised learning framework (we can just define everything inside the first sum as $\nabla g(\tau)$).}

@@

\section{Algorithm}

@@

\begin{itemize}
	\item The objective $J(\vtheta)$ is typically non-convex;
	\item The dataset $\mathcal{T}$ can be infinite;
	\item The value of the parameter $\vtheta$ influences the sampling of trajectories;
\end{itemize}

@@

\section{Convergence Guarantees}
\begin{assumption}\label{ass:bounded_score}
For each trajectory $\tau$, value of $\vtheta$ and dimensions $i,j$ there are positive constants $G,F<\infty$ such that:
\begin{align*}
	&\left|\nabla_{\theta_i}\log\pi_{\vtheta}(\tau)\right| \leq \GRADLOG \\
	&\left|\frac{\partial^2}{\partial\theta_i\partial\theta_j}\log\pi_{\vtheta}(\tau)\right| \leq \HESSLOG
\end{align*}
\end{assumption}

\begin{assumption}\label{ass:M2}
	There is a constant $W<\infty$ such that, for each pair of policies encountered in Algorithm \ref{alg:svrpg} and for each possible trajectory,
\[
	\mathbb{V}ar\left(\omega(\tau)\right) \leq \VARIS
\]
\end{assumption}

\begin{assumption}\label{ass:REINFORCE}
There is a constant $V<\infty$ such that, for each policy $\pol$:
\[
	\Tr\mathbb{V}ar\left(\gradApp{\vtheta}{1}\right) \leq \VARRF
\]
\end{assumption}

\begin{restatable}[]{lemma}{L-smoothness}\label{lemma:lsmooth}
Under Assumption \ref{ass:bounded_score}, $J(\vtheta)$ is L-smooth.
\end{restatable}

\begin{restatable}[Convergence of the SVRPG algorithm]{theorem}{convergence}\label{theo:convergence}
Under Assumptions \ref{ass:bounded_score} and \ref{ass:M2}, the parameter vector $\vtheta_A$ returned by Algorithm \ref{alg:svrpg} after $T=m\times S$ iterations has, for some positive constants $\psi,\zeta, \xi$ and for proper choice of the step size, the following property:
\begin{align*}
	&\EVV[]
	{\norm[]{\nabla J(\vtheta_A)}^2} 
		\leq
		\frac{J(\vtheta^*)-J(\vtheta_0)}{\psi T} +
		\frac{\zeta}{N}
		+\frac{\xi}{B}
\end{align*}
\end{restatable}

@@

\section{Experiments}
We have designed our experiments to investigate whether:
\begin{enumerate}
\item The algorithm reaches high value of the reward function faster than GPOMDP.
\item Averagely SVRPG keep the reward function more stable than GPOMDP.
\end{enumerate}

@@

\subsection*{Ancillary Lemmas}
Before addressing the main convergence theorem, we prove some useful lemmas.

@@

\section{Experimental Details}\label{app:proofs}
Setting specifications for all the used environments:
\begin{enumerate}
	\item \emph{Cart-Pole Balancing} : 4-dimensional state space: cart position x, pole angle $\theta$, cart velocity $\dot{x}$ and pole velocity $\dot{\theta}$; 1-dimensional action space: the horizontal force applied to the cart body. Reward function  is defined as $r(s, a) := 10 - (1 - cos(\theta)) - 10^{-5}\norm[] a^2$. The episodes terminates when $|x|>2.4$ or $|\theta|>0.2$ or the number of time steps T is greater than 100.
	\item \emph{Mujoco Swimmer}: 12-dimensional state space: 3 links velocities ($v_x$ and $v_y$ of center of masses) and 2 actuated joints angles. 2-dimensional action space: the two momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 10^{-4}\norm[2] a^2$. The episodes terminates when the number of time steps T is greater than 500.
	\item \emph{Mujoco Half Cheetah}: 20-dimensional state space: 9 links and 6 actuated joints angles. 6-dimensional action space: the 6 momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 0.05\norm[2] a^2$. The episodes terminates when the number of time steps T is greater than 500.
\end{enumerate}
